---
title: "Seurat PBMC Tutorial"
output:
  html_document:
    theme: united
    df_print: kable
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
# Hide extraneous output in compiled document
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

# Setup the Seurat Object

```{r init}
library(dplyr)
library(Seurat)
library(patchwork)

pbmc.data <- Read10X(data.dir = "./raw_data/pbmc3k/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc

```

## Peek at data in a .mtx file

```{r}
pbmc.data[c("CD3D","TCL1A","MS4A1"), 1:30]

# Size as dense matrix
dense.matrix <- as.matrix(pbmc.data)
dense.matrix.size = format(object.size(dense.matrix), units="MB")
sprintf("Size as dense matrix: %s", dense.matrix.size)

# Size as sparse matrix
sparse.matrix <- as(dense.matrix, "dgCMatrix")
sparse.matrix.size <- format(object.size(sparse.matrix), units="MB")
sprintf("Size as sparse matrix: %s", sparse.matrix.size)

```

## Quality Control
Low-quality / dying cells often exhibit extensive mitochondrial contamination.
We calculate mitochondrial QC metrics with the PercentageFeatureSet() function, which calculates the percentage of counts originating from a set of features.
We use the set of all genes starting with MT- as a set of mitochondrial genes
Note: It seems that "features" means "genes" here?

```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- Seurat::PercentageFeatureSet(pbmc, pattern = "^MT-")
```


In the example below, we visualize number of features, total molecule count, and percetage of features that are mitochondrial

```{r}
# Visualize QC metrics as violin plots
VlnPlot(pbmc, features=c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

We use these visualizations to filter cells.

We filter cells that have unique feature counts over 2,500 or less than 200
- Cell doublets or multiplets may exhibit an aberrantly high gene count
- Low quality cells or empty droplets will often have very few genes
We filter cells that have >5% mitochondrial counts






